---
title: "Datawrangling"
author: "Luke Singh"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---





Some of the exercises in this lab are based on those found in  [Wickham and Grolemund, R for Data Science](https://r4ds.had.co.nz/).

 

## Getting started

(@)  Open R Studio and open the STAT702 R project that you created last week. 

(@) Download "STAT702_week03_lab_datawrangling2_answer.Rmd" from Blackboard and save in your project directory.

(@) Knit the file. You will write your answers to the following questions in that file.


(@) In this lab we will need a number of packages. 

```{r requiredPackages}
# List of packages
packages <- c("tidyverse", "readxl", "nycflights13", "lubridate")
```

- Install them, if required, by uncommenting the following code.
```{r installPackages}
# Find packages that are not installed
packages_to_install <- packages[!(packages %in% installed.packages()[,"Package"])]

# Install required packages
#if(length(packages_to_install) > 0) install.packages(packages_to_install, 
#                                                     dependencies=TRUE))
```

- Load the packages
```{r loadPackages}
# Load new packages
sapply(packages, library, character.only=TRUE)
```


```{block, box.title = "Tip", box.body = list(fill = "lightblue")}
You can get help in R by typing "?" followed by the function name. For example ?ggplot
```

## Joins


(@) Consider the following data set.

  - Run the following code and inspect the outputs.
```{r inspect_tibbles, results="hide"}

STAT702exams <- tribble(
  ~studID, ~grade,
     1, "A",
     2, "B",
     3, "C+"
)
STAT702classlist <- tribble(
  ~studentID, ~name,
     1, "Charlotte",
     2, "Zoe",
     3, "Caitlin",
     4, "Abel"
)


STAT700classlist <- tribble(
  ~studentID, ~name,
     1, "Charlotte",
     2, "Zoe",
     6, "Conor",
     7, "Archie"
)

```

Use the appropriate \verb|join| function to create a dataset that can answers the following questions.

- Create a dataset containing the name, ID and exam mark of all STAT702 students. Include students who didn't sit the exam.

```{r}
ggplot(data = STAT700classlist) +
  geom_point(mapping = aes(x = names, y = studentID), square = TRUE)
```



- Create a dataset containing the name, ID and exam mark of the STAT702 students who sat the exam.  

```{r}
STAT700classlist %>%
  right_join(STAT702exams, by=c("studentID"="studID"))
```


- Create a dataset containing 
the name and ID of the STAT702 students who sat the exam.

```{r}
STAT700classlist %>%
  semi_join(STAT702exams, by=c("studentID"="studID"))
```



- Which students didn't sit the STAT702 exam?


```{r}
STAT700classlist %>%
  anti_join(STAT702exams, by=c("studentID"="studID"))
```


- Create a dataset containing the name and ID of all students studying either of the level 7 STAT papers.
Ensure your new dataset has only 2 columns: one for ID and one for name.

```{r}
STAT700classlist %>%
  full_join(STAT702classlist, by=c("studentID" = "studentID", "name" = "name"))
```



- Create a dataset containing the name and ID of all students studying both of the level 7 STAT papers.
Ensure your new dataset has only 2 columns: one for ID and one for name.

```{r}
STAT700classlist %>%
  inner_join(STAT702classlist, by=c("studentID" = "studentID", "name" = "name"))
#semi_join works too but inner join will allow dupes if any.
```


(@)  Use the appropriate \verb|join| function to create a dataset that can answer the following questions.
You will need to you some other functions well (e.g. \verb|filter|).

- Create a table showing the names of the destinations that JetBlue Airways flies to.  Include a column showing the number of flights to each destination.
`

Start from this code:
```{r jetblue_starter}
selected_airlines <- airlines %>% filter(name %in% c("JetBlue Airways"))
```

 
```{r}
jetblue <- flights %>%
  semi_join(selected_airlines, by = "carrier") %>%
  count(dest) %>%
  left_join(airports, by = c("dest" = "faa")) %>%
  select(dest, name, n)
```

 
- What destinations do both JetBlue Airways and United Air Lines Inc (UA) fly to?
 
```{r}
united <- flights %>%
  filter(carrier == "UA") %>%
  count(dest) %>%
  left_join(airports, by = c("dest" = "faa")) %>%
  select(dest, name, n)

inner_join(jetblue, united, by = c("dest" = "dest", "name" = "name"),
           suffix = c("JetBlue", "United"))
```


(@) Use ggplot to construct a bar plot showing the number of flights per month. Add a title and subtitle to your plot.
```{r}
flights %>% group_by(month) %>%
  summarize(Number_of_Flights = n ()) %>%
  ggplot()+
  geom_bar(mapping = aes(x = month, y = Number_of_Flights), stat = "identity",
           fill = "darkblue")+
  labs(title = "Monthly Total Flights",
       subtitle = paste("2013"),
       y = "Number of Flights",
       x = "Time") + theme_bw(base_size = 20) +
  scale_x_continuous("Month", breaks = 1:12, labels = month.abb)
```

 
 
 
## Strings

(@) Use \verb|str_view| and \verb|str_view_all| to identify regular expressions which do the following:

- Identify all digits
```{r str_view_setup}
x <- c("25th May 2019", "April 19th 2005", "12 December 1984")
```


```{r}
# Write your answer here
```

 
- Identify the year in each string. 

Hint. Consider the following code.
```{r str_view_first2digits}
str_view_all(x, "^[\\d]{2}")
```
 

```{r}
str_view_all(x, "[\\d]{4}$")
```

 
- Identify the month
```{r}
str_view_all(x, "[A-Z][a-z]*")
```


(@)  Use the appropriate \verb|stringr| function for each of the following questions.

- Sort in alphabetical order

```{r sort}
x <- c("apple", "pear", "banana", "eggplant")
```

```{r}
str_sort(x)
```


- Replace all vowels with  \verb|-|.

```{r vowels}
x <- c("apple", "pear", "banana", "eggplant")
```

```{r}
str_replace_all(x, "[aeiou]", "-")
```

 
- Combine the following vectors to create a vector containing the day, month and year, e.g. "12 September 2010".
```{r strc}
day <- c(1, 2, 3, 4)
month <- c("April", "May", "May", "July")
year <- c(2009, 2019, 1999, 2005)
```

```{r}
str_c(day, month, year, "")
```



(@)  Using the flights data, create a new tibbble which contains only following sentences about each flight.

Hint: consider the functions \verb|transmute| and \verb|str_c|.

Test your code on this subset of data.
```{r challenge_setup}
flights_subset <- flights[1:10,]
```

 
- Flight UA1545 from EWR to IAH on 1/1/2013, departed at 517.
```{r}
flights_subset %>% transmute(
  description = str_c("Flight ", flight, "from ", origin, "to ", dest, "on ", day, "/", month, "/", year,
                      ", departed at ", dep_time, ".", sep= "")
)
```



- Challenge: Flight UA1545 from EWR to IAH on 1 January 2013, departed 2 minutes late at 5:17am.

## Dates and Times

(@) Get started

- Load the packages (if you haven't done so already)
```{r}
#install.packages("lubridate")
#library(lubridate)
#library(nycflights13)
```
- Check package has installed. Find out what the date and time is, right now.

```{r}
# this is a date
today()

# this is a date-time
now()
```

(@) Creating dates and times

- Run the following code to create dates from strings. Inspect the output, then modify the code to be today's date.
```{r}
ymd("2017-01-31")
mdy("January 31st, 2017")
dmy("31-Jan-2017")
```

- Run the following code to create date-times from strings. Inspect the output, then modify the code to be today's date.
```{r}
ymd_hms("2017-01-31 20:11:59")
mdy_hm("01/31/2017 08:01")
```

- Dates and date-times can also be created from individual variables. Create a `departure` variable using the year, month, day, hour and minute in the flights dataset.

```{r}
flights %>% 
  select(year, month, day, hour, minute)

flights %>% 
  select(year, month, day, hour, minute) %>% 
  mutate(departure = make_datetime(year, month, day, hour, minute))

```


- Using the variables `year`, `month`, `day` and  `dep_time`, create a new date-time variable called a `departure_time`. Use the function `select` to make the new departure time variable the first column,  followed by all other variables. 
Save the resulting tibble as `flights_dt`.

Hint: consider how you can use modulus arithmetic to extract the relevant information from `dep_time`.

```{r}
123 %% 100
123 %/% 100
```


```{r}
flights_dt <- flights %>%
  filter(!is.na(dep_time)) %>%
  mutate(departure_time = make_datetime(year, month, day, dep_time %/%100,
                                        dep_time %% 100)) %>%
  select(departure_time, everything())
```


- Visualise the departure times on 1st January 2013 using the `flights_dt` dataset.



(@) Extracting date-time components 

- Use the function `ymd_hms` to create a date-time object with the current date and time. Save the object as a variable called `datetime`.

```{r}
flights_dt %>%
  filter(departure_time < ymd(20130102)) %>%
  ggplot(aes(departure_time)) +
  geom_freqpoly(bindwidth = 600)
```


- Use the function `year` to extract the value of the year from your date-time object.



- Use the function `month` to extract the value of the month from your date-time object.   

```{r}
datetime <- ymd_hms("2021-03-19 12:01:56")
```


- Use the function `month` to extract the value of the month from your date-time object.  What do the options `label` and `abbr` do? Use `?month` to find out more, then test your answer by running the code in R. 

```{r}
month(datetime, label = TRUE, abbr = FALSE)
```


- What do the functions `mday`, `yday` and `wday` do?
```{r}
mday(datetime)
yday(datetime)
wday(datetime)
```


- Using the tibble `flights_dt` create a new variable called `month2` by extracting the month from the `departure_time` column.
Use the function select to select only the `departure_time` and `month2` columns. Check that `month2` correctly extracts the month.

```{r}
flights_dt %>% mutate(
  month2= month(departure_time)
) %>% select(departure_time, month2)
```


## Further practice

(@)  Review the examples in the lecture notes
(@)  Read R4DS chapters 13 - 16, 27 and do the exercises.

