---
title: 'STAT702: Week 1 Lab'
output:
  unilur::tutorial_pdf: default
  unilur::tutorial_pdf_solution:
    md_extensions: +example_lists
  unilur::answer_rmd: default
---

```{r setup, echo=FALSE, include = FALSE, results = "hide", message=FALSE}
library(knitr)
figprefix = paste('Rfigs/', "STAT702_week01_", sep="")
opts_chunk$set(fig.path=figprefix,fig.align='center', fig.show='hold',  out.width="50%")

str(knitr::opts_template$get("solution"), give.attr = FALSE)
knitr::opts_template$set(solution = list(box.title = "Solution",
                box.body = list(fill = "lightcyan2", color="black"),
                box.header = list(fill = "lightcyan2", colour="black"), 
                box.collapse = TRUE))

#install.packages("remotes")
#remotes::install_github("koncina/unilur")

```

`````{block, solution=TRUE}
This file contains the solutions.
```

Some of the exercises in this lab are based on those found in  [Wickham and Grolemund, R for Data Science](https://r4ds.had.co.nz/).



#### Getting started
(@) Install the `tidyverse` package

```{r install_tidy, eval=FALSE}
install.packages("tidyverse")
```

(@) Load the `tidyverse` package
```{r load_tidy}
library(tidyverse)
```


```{block, box.title = "Tip", box.body = list(fill = "lightblue")}
You can get help in R by typing "?" followed by the function name. For example ?ggplot
```

 
#### ggplot2

(@) Inspect the `mpg` dataset using the following code and answer the following questions.

```{r inspect_mpg, results="hide"}
mpg
```

  - How many cars are included in this dataset?
```{r solution=TRUE}
#multiple ways to get this answer
dim(mpg)
nrow(mpg)
```


  - What does the variable "cty" represent?
```{r, eval=FALSE, solution=TRUE}
?mpg
#cty = city miles per gallon
```


(@) Do cars with big engines use more fuel than cars with small engines?  Use `ggplot` to create a plot to explore this question.

```{r, solution = TRUE}
ggplot(data = mpg) +
  geom_point(mapping=aes(x=displ, y=hwy))
```

(@) Add a colour aesthetic to the plot to represent the type of transmission.

```{r, solution = TRUE}
ggplot(data = mpg) +
  geom_point(mapping=aes(x=displ, y=hwy, col=trans))
```

(@) Does the fuel efficiency differ for cars of different classes? Create a boxplot, with one box for each class.


```{r, solution = TRUE}
ggplot(data = mpg) +
  geom_boxplot(mapping=aes(y=hwy, x=class)) 
```

(@) In the previous boxplot, flip the axes.


```{r, solution = TRUE}
ggplot(data = mpg) +
  geom_boxplot(mapping=aes(y=hwy, x=class)) + 
  coord_flip()
```


(@) In the following graph, why are the points not blue?

```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ, y=hwy, col="blue"))
```

`````{block, solution = TRUE}
They are not blue because the command is inside the aes() function. It should be outside it.
```
 
 
 
#### dplyr

```{r}
library(dplyr)
```

(@) Load and inspect the `flights` dataset

```{r}
library(nycflights13)
flights
```

(@) What does the variable "air_time" represent?

```{r, solution=TRUE, eval=FALSE}
?flights
#air_time = Amount of time spent in the air, in minutes
```

(@) How many flights departed on 1st March 2013?

```{r, solution=TRUE}
march1 <- filter(flights, month==3, day==1) 
nrow(march1)
#958 flights
```

(@) For the flights which departed on 1st March 2013, what was the destination of the one with the biggest departure delay?

```{r, solution=TRUE}
# There are multiple ways to write the code for this question. This is one of them.

# sort flights be departure time
arrange(march1, desc(dep_delay)) %>%
# select relevant columns (not required, but makes it clearer)
  select(dest, dep_delay) %>%
# extract top entry
  top_n(n=1)  -> late_dep

late_dep

# Use airport metadata to get name of airport
filter(airports, faa==late_dep$dest)

# Dallas Fort Worth Intl
```


(@) Which destination has the greatest mean departure delay in 2013? Hint: use group_by and summarise

```{r, solution=TRUE}
#  South Bend Rgnl
flights %>% 
  group_by(dest) %>% 
  summarise(mean_dep_delay = mean(dep_delay, na.rm = TRUE)) %>%
  arrange(desc(mean_dep_delay)) %>%
  slice(1) -> dep_delay_max_average

filter(airports, faa==dep_delay_max_average$dest)$name
```

(@) Use summarise to determine the number of flights by each carrier.

Hint: `summarise(n=n())`
```{r, solution=TRUE}
flights %>% group_by(carrier) %>% summarise(n=n())

```

(@) For flights which departed on 1st March, create a boxplot showing the departure delay, with one box for each "carrier". Which airline seems to have the worst on-time performance?

```{r, solution=TRUE}
flights %>% 
  filter(month==3, day==1) %>%
ggplot() + 
  geom_boxplot(mapping = aes(y=dep_delay, x=carrier))
```


#### Application

Pick one of the following datasets:

* `ggplot2::mpg`
* `nycflights13::flights`
* `ggplot2::diamonds`

(@) Look at the help page in R for your chosen dataset, e.g. `?flights`.

(@) In groups of 2 or 3, write 3 questions about this dataset.

e.g. Is there a relationship between engine size and fuel efficiency?

(@) Create an appropriate graph or summary statistics to investigate each of your questions.


#### Further practice

(@)  Read R4DS chapters 1, 2, 3 and 5 and do the exercises. If you are new to R, read chapter 4 too.

